---
title: "Advanced Plotting: Geographical, Longitudinal, and Time Series"
format: 
  live-html:
    toc: true
resources: 
  - "Data-sets"
  
execute:
  echo: true
  warning: false
  message: false

webr:
  editor: true
  render-df: gt-interactive
    
embed-resources: true

filters:
  - custom-numbered-blocks

custom-numbered-blocks:
  classes:
    Assumption:
      colors: [FFCDD2, F44336]
      boxstyle: foldbox.simple
      collapse: false
    Example:
      colors: [BBDEFB, 2196F3]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: [C8E6C9, 4CAF50]
      boxstyle: foldbox.simple
      collapse: false
    Technical-point:
      colors: [FFF9C4, FFEB3B]
      boxstyle: foldbox.simple
      collapse: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

While we have covered the most integral and widely used plots, there are still many more to explore depending on the type of data you have.

In this chapter, we will briefly go through examples of some advanced plotting and visualisation you can do in R.

# Geographical Data

There are many nice packages out there to represent geographical data. One of the most popular packages is [`leaflet`](https://rstudio.github.io/leaflet/), which is an open-source JavaScript library for interactive maps.

```{r, message = FALSE}
library(leaflet)
library(tidyverse)
m = leaflet() %>% addTiles()
m  

m = m %>% setView(151.64261320531202, -30.48717288612027, zoom = 17)

m %>% addPopups(
  151.64261320531202,
  -30.48717288612027,
  'Here is the <b>Mathematics, Statistics and Computer Science Building</b>, UNE'
)

```

You can customise the map even further:\

```{r}
circles <- data.frame(
  lng = c(151.643),
  lat = c(-30.487)
)

leaflet() %>%
  addTiles() %>%
  setView(lng = 151.643, lat = -30.487, zoom = 14) %>%
  addCircleMarkers(data = circles, color = "red") %>%
  addCircles(data = circles, radius = 1500)

```

You can read more about other customisations here. https://r-charts.com/spatial/interactive-maps-leaflet/

Want to learn more about creating maps in R? check out these resources:

# Longitudinal

# Times Series
